---
title: "Trends & Variations in Transportation Demand in Chicago"
author: "Amir Kazi"
date: "1/27/2019"
output:
  html_document: 
    code_folding: hide
---


```{r setup, include=FALSE}
library (tidyverse)
library (here)
library(lubridate)  # lubridate masks object 'here' from package 'here'.
library(scales)
library(ggrepel)
```


```{r}
# READING DATA

bus_trips <- read.csv (here::here("data", "bus_data.csv"))
train_trips <- read.csv (here::here("data", "train_data.csv"))
weather_descriptions <- read.csv(here::here("data", "weather.csv"))
weather <- read.csv(here::here("data", "midway_weather.csv")) 
  
  
```

```{r}

train_trips$date <- ymd(train_trips$date)
train_trips$dayofweek <- wday(train_trips$date, label=TRUE)  
train_trips$day <- day(train_trips$date)
train_trips$month <- month(train_trips$date, label = TRUE)
train_trips$year <- year(train_trips$date)

bus_trips$date <- ymd(bus_trips$date)
bus_trips$dayofweek <- wday(bus_trips$date, label=TRUE)  
bus_trips$day <- day(bus_trips$date)
bus_trips$month <- month(bus_trips$date, label = TRUE)
bus_trips$year <- year(bus_trips$date)

train_trips$trip_identity = 'train'
bus_trips$trip_identity = 'bus'

train_trips$route = NA
bus_trips$station_id = NA
bus_trips$stationname = NA

public_transport = rbind(train_trips,bus_trips)
public_transport = filter (public_transport, year(date) != 2018 )

public_transport$dayofweek <- factor(public_transport$dayofweek, levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun"))

weather$date <- ymd(weather$DATE)
weather$dayofweek <- wday(weather$DATE, label=TRUE)  
weather$day <- day(weather$DATE)
weather$month <- month(weather$DATE, label = TRUE)
weather$year <- year(weather$DATE)
weather$dayofyear <- yday(weather$DATE)

```


```{r, warning=FALSE, message=FALSE}

# PLOT 1
# 3 Year Chicago Public Transportation Trend + Weather

# Shading Winters From 21st December to 30th March
# Reference: https://stackoverflow.com/questions/18419628/creating-custom-legends-in-ggplot2


df<-data.frame(xmin=as.Date(c('2015-01-01','2015-12-21','2016-12-21', '2017-12-21' )),
               xmax=as.Date(c('2015-03-20','2016-03-30', '2017-03-30', '2017-12-31' )),
               ymin=c(-Inf,-Inf),
               ymax=c(Inf,Inf),
               winters=c("Winters"))


public_transport %>%
  group_by(date, trip_identity) %>%
  summarise (count = sum(rides)) %>%
  ggplot(aes(x = date, y = count/1000)) +
    geom_smooth(aes(color = trip_identity), se = FALSE, na.rm = TRUE) +
    geom_rect(data=df,aes(xmin=xmin,ymin=ymin,xmax=xmax,ymax=ymax,fill=winters),
                      alpha=0.15,inherit.aes=FALSE)+
    scale_fill_manual(values=c("#FF66CC")) + 
    labs(title ="Daily Demand for Public Transport in Chicago is lowest in Winter", 
         subtitle = "Train usage is more elastic to seasonal changes than bus usage.",
         x = "", y = "Passengers Per Day (Thousands)",
         caption = "CTA Ridership Data 2015-2017") +
    theme_bw() +
    theme(plot.title = element_text(hjust = 0, size = 12, face = "bold"), 
          plot.subtitle = element_text(size = 10, face = "bold"),
          legend.title=element_blank(),
          legend.spacing.x = unit(0.2, 'cm'))  + scale_x_date(date_labels = "%d %b %Y", 
                                                              date_breaks = "1 year", 
                                                              date_minor_breaks = "1 month")+
    scale_y_continuous(labels = scales::comma) +
  geom_vline(xintercept = as.Date(c("2015-01-01", "2016-01-01", "2017-01-01", "2018-01-01"))
             , linetype = 3) 
 
```


Between 2015 and 2017, it is apparent that there is a variation in the demand
for buses and trains throughout the year, peaking in summer time, and plunging
in the winter. However, while trains experience a roughly similar pattern each year,
the demand for buses seems to be falling across years. These patterns can impact
the allocation and deployment of transportation services by the CTA across seasons.


```{r, warning=FALSE, message=FALSE}
# reference: https://ibecav.github.io/slopegraph/

p <- public_transport %>%
  group_by(year, month, trip_identity) %>%
  summarise(riders = sum(rides)) %>%
  filter (trip_identity == "bus") 

  ggplot(p, aes(x = year, y = riders, group = month)) +
  geom_line(aes(color = month), alpha = 0.3, size = 2) +
  geom_point(aes(color = month, alpha = 1), size = 3) +
  geom_text_repel(data = p %>% filter(year == "2015"),
                  aes(label = paste0(month)) , 
                  hjust = "left", 
                  fontface = 7, 
                  size = 4, 
                  nudge_x = -.45, 
                  direction = "y") +
  geom_text_repel(data = p %>% filter(year == "2017"),
              aes(label = paste0(month)) , 
              hjust = "right", 
              fontface = 7, 
              size = 4, 
              nudge_x = .5, 
              direction = "y") +
  scale_x_discrete(position = "bottom") +
  theme_bw() +
  theme(legend.position = "none") + 
    annotate("text", x = 2015, y = 25500000, label = "2015", family = "Courier")  + 
    annotate("text", x = 2016, y = 25500000, label = "2016", family = "Courier") +  
    annotate("text", x = 2017, y = 25500000, label = "2017", family = "Courier") +
  theme(panel.border     = element_blank()) +
  theme(axis.title.x     = element_blank()) +
  theme(axis.text.y     = element_blank()) +
  theme(panel.grid.major.x = element_blank()) +
  theme(axis.ticks       = element_blank()) +
  labs(
    title = "Overall Bus Usage Has Declined Between 2015 and 2017",
    subtitle = "Summer Months still experience more bus usage than in Winter",
    caption = "CTA Transit Data 2015 - 2017",
    y = "Total Bus Riders") +
  theme(plot.title = element_text(hjust = 0.58, size = 12, face = "bold"), 
        plot.subtitle = element_text(hjust = 0.515, size = 10, face = "bold")) + scale_y_continuous(labels = scales::comma)

```

Comparing bus usage across different months between 2015 - 2017, it can be seen that
total bus usage has declined for each month across these periods. However, relatively,
demand is still higher in summer months than in winter months. The overline decline may be
due to a shift towards other forms of transportation such as taxis and other ride-sharing vehicles.


```{r, warning=FALSE, message=FALSE}
# PLOT 2
# Weekly Demand of Public Transportation

# Variation in usage by day of week
# 3*52.19 == Number of weeks in 3 years


public_transport %>%
  group_by(dayofweek, trip_identity) %>%
  summarise(count = (sum(rides)/(3*52.19))/1000)  %>%
  ggplot(aes(x=dayofweek, y=count))  +
  geom_point(aes(col=trip_identity), size = 5, stroke = 1) + 
  geom_line(mapping = aes(x = dayofweek, y=count),  alpha = 0.3)  +
  scale_y_continuous(labels = scales::comma) +
  coord_cartesian( ylim = c(200, 900)) + 
  labs(title ="Usage of Public Transport in Chicago is Highest on Weekdays", 
       subtitle = "Weekends see a significant drop in riders",
       x = "Day of Week", y = "Average Weekly Passengers (Thousands)",
       caption = "CTA Ridership Data 2015-2017") +
  theme_bw()  + 
  theme(plot.title = element_text(hjust = -0, size = 12, face = "bold"), 
        plot.subtitle = element_text(size = 10, face = "bold"), 
        legend.title=element_blank(),
        legend.key = element_rect(fill = "white")) +
  coord_flip()

```

Comparisons within a week indicate that highest usage is during weekdays, possibly due to work or school transits. Deployments of public transportation are also more spread out on weekends by the CTA to ensure more efficient use.


```{r, warning=FALSE, message=FALSE}

# PLOT 3
# Physical Manifestations of Seasonal Changes
# SNOW & PRECIPITATION LEVELS CHANGE

weather %>%
  group_by(month) %>%
  summarize(avg_prcp = mean(PRCP, na.rm = TRUE), avg_snow = mean(SNOW, na.rm = TRUE)) %>%
  ggplot() +
  geom_step(aes(x = month, y = avg_prcp, group = 1), linetype = 14, colour = "cyan4") +
  geom_step(aes(x = month, y = avg_snow, group = 1), linetype = 11, colour = "indianred1") +
  theme_bw() + 
  annotate(geom="text", x=3.55, y=7, label="SNOW", color="indianred3", fontface =4) +
  annotate(geom="text", x=7, y=4.9, label="PRECIPITATION", color="cyan4", fontface =4) +
  labs(title ="Larger Seasonal Variation in Snow than in Precipitation in Chicago", 
       subtitle = "Precipitation Slightly Higher During Summer ",
       x = "", 
       y = "Monthly Average (Inches)",
       caption = "National Centers for Environmental Information 2015-2017") +
  theme(plot.title = element_text(hjust = -0, size = 12, face = "bold"), 
        plot.subtitle = element_text(size = 10, face = "bold"))

```

Winter alone doesn't cause a decline in demand - there is a very real effect due to an increased accumulation of snow in winter months which may make it more inconvenient to take public transportation, especially trains, if it requires walking large distances to a train station.



```{r, warning=FALSE, message=FALSE}
# PLOT 4
# SNOW DAY - AND AFTER

snow_days <-
  weather %>% 
  filter (SNOW > 0)

day_after_snow <-
    snow_days %>%
    select(date) + 1
  
train_snow_days <- 
  merge(snow_days, train_trips, by = "date") %>%
  group_by(date) %>%
  summarize (riders = sum(rides)) %>%
  mutate (type = "TRAIN", day_type = "SNOW")

train_after_snow <- 
  merge(day_after_snow, train_trips, by = "date") %>%
  group_by(date) %>%
  summarize (riders = sum(rides)) %>%
  mutate (type = "TRAIN", day_type = "DAY_AFTER_SNOW")

bus_snow_days <- 
  merge(snow_days, bus_trips, by = "date") %>%
  group_by(date) %>%
  summarize (riders = sum(rides)) %>%
  mutate (type = "BUS", day_type = "SNOW")

bus_after_snow <- 
  merge(day_after_snow, bus_trips, by = "date") %>%
  group_by(date) %>%
  summarize (riders = sum(rides)) %>%
  mutate (type = "BUS", day_type = "DAY_AFTER_SNOW")


snow_comparison <- do.call("rbind", list(train_snow_days, train_after_snow, bus_snow_days,bus_after_snow ))
snow_comparison$day_type <- factor(snow_comparison$day_type, levels = c("SNOW", "DAY_AFTER_SNOW"))

ggplot(snow_comparison) +
  geom_violin(aes(x = day_type, y = riders/1000,  fill = type)) + 
  scale_x_discrete(labels = c('Rides Taken on Snow Day','Rides Taken on Day After Snow Day')) +
  labs(title ="Fewer Public Transportation Riders on Day After Snow", 
       subtitle = "Not all days experience a decline, however.",
       x = "", 
       y = "Riders (Thousands)",
       caption = "NCEI & CTA Rider Data 2015-2017") +
  theme_bw() +
  theme(plot.title = element_text(hjust = -0, size = 12, face = "bold"), 
        plot.subtitle = element_text(size = 10, face = "bold"),
        legend.title=element_blank(),
        legend.key = element_rect(fill = "white"),
        legend.spacing.x = unit(0.2, 'cm'))




```


For rides taken on the day after a snow day, it can be seen that there is an overall decrease in ridership throughout different ridership levels for buses and trains. This may indicate a preference to stay indoors, to use personal vehicles, or to shift towards taxis or ride-share options.


Further research will assess impact of weather on taxi and divvy demand, as well as variations in demand resulting from other factors such as Chicago Bears/Cubs games, concerts/conferences, and large-scale events (such as Lolapalooza).


