---
title: "Trends & Variations in Transportation Demand in Chicago"
author: "Amir Kazi"
date: "February 2019"
output:
  html_document: 
    code_folding: hide
---

```{r setup, include=FALSE}
library (tidyverse)
library (here)
library(scales)
library(ggrepel)
library(gtools)
library(forcats)
library(sf)
library(png)
library(grid)
library(lubridate)  # lubridate masks object 'here' from package 'here'.
```

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=10, fig.height=8, echo=FALSE, warning=FALSE, message=FALSE)
```


```{r, warning=FALSE, message=FALSE, include=FALSE}
# READING PUBLIC TRANSPORTATION DATA
bus_trips <- read.csv (here::here("data", "bus_data.csv"))
train_trips <- read.csv (here::here("data", "train_data.csv"))

# READING WEATHER DATA
#weather_descriptions <- read.csv(here::here("data", "weather.csv"))
weather <- read.csv(here::here("data", "midway_weather.csv")) 

# READING DIVVY DATA
divvy_trips <- read.csv (here::here("data", "Divvy_Trips.csv"))
### columns to drop from divvy
### TRIP.ID, BIKE.ID, USER
#divvy_trips <-select(divvy_trips, -c()) 
#select (mydata,-c(mpg,cyl,wt))

# CREATING PUBLIC TRANSPORTATION DATAFRAME
train_trips$date <- ymd(train_trips$date)
train_trips$dayofweek <- lubridate::wday(train_trips$date, label=TRUE)  
train_trips$day <- day(train_trips$date)
train_trips$month <- lubridate::month(train_trips$date, label = TRUE)
train_trips$year <- year(train_trips$date)

bus_trips$date <- ymd(bus_trips$date)
bus_trips$dayofweek <- lubridate::wday(bus_trips$date, label=TRUE)  
bus_trips$day <- day(bus_trips$date)
bus_trips$month <- lubridate::month(bus_trips$date, label = TRUE)
bus_trips$year <- year(bus_trips$date)

train_trips$trip_identity = 'train'
bus_trips$trip_identity = 'bus'

train_trips$route = NA
bus_trips$station_id = NA
bus_trips$stationname = NA

public_transport = rbind(train_trips,bus_trips)
public_transport = filter (public_transport, year(date) != 2018 )

public_transport$dayofweek <- factor(public_transport$dayofweek, levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun"))

# DIVVY DEMAND GROUPED BY DAY
divvy_trips$date <- mdy_hms(divvy_trips$START.TIME)

divvy_trips$time_of_day <- hour(divvy_trips$date)
divvy_trips$day_of_week <- wday(divvy_trips$date, label = TRUE)

# SETTING WEATHER DATAFRAME
weather$date <- ymd(weather$DATE)
weather$dayofweek <- lubridate::wday(weather$DATE, label=TRUE)  
weather$day <- day(weather$DATE)
weather$month <- lubridate::month(weather$DATE, label = TRUE)
weather$year <- year(weather$DATE)
weather$dayofyear <- yday(weather$DATE)
```


```{r, warning=FALSE, message=FALSE, include=FALSE}
# MY THEME

my_theme <- 
  theme(
    axis.text = element_text(size = 7, face = "bold", family="Courier"),
    axis.title =  element_text(size = 10, face = "bold", family="Courier"),
    plot.title = element_text(hjust = 0, size = 14, face = "bold", family="Courier"), 
    plot.subtitle = element_text(size = 12, face = "bold", family="Courier"),
    plot.caption = element_text(size = 8, face = "bold", family="Courier"),
    legend.title=element_blank(),
    legend.spacing.x = unit(0.2, 'cm'),
    panel.border = element_rect(colour = "#969696", fill=NA, size=2),
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank())
    

# CHOOSING COLOURS FOR TYPE OF TRANSPORTATION
taxi_color <- "#fec44f"  #yellow
divvy_color <- "#6baed6" #blue
bus_color <- "#1B9969" #green
train_color <- "#c51b8a" #violet
winter_color <- "#7a859f"
divvy_female_color <- "#2171b5"
divvy_male_color <- "#6baed6"
```



```{r, warning=FALSE, message=FALSE, include=FALSE}
# GETTING TAXI START TIMESTAMPS
taxi_2015 <- read.csv (here::here("data", "2015_Taxi_Times.csv"))[,1]
taxi_2016 <- read.csv (here::here("data", "2016_Taxi_Times.csv"))[,1]
taxi_2017 <- read.csv (here::here("data", "2017_Taxi_Times.csv"))[,1]

taxi_start_times <- mdy_hms(fct_c(taxi_2015, taxi_2016, taxi_2017))
taxi_start_times_date <- as.data.frame(date(taxi_start_times))

colnames(taxi_start_times_date) <- c("date")
remove(taxi_2015, taxi_2016, taxi_2017)

# PUBLIC TRANSPORTATION DEMAND GROUPED BY DAY
public_transport_by_day <-
  public_transport %>%
    group_by(date, trip_identity) %>%
    summarise(riders = sum(rides)) %>%
    select(date, riders, trip_identity)

divvy_by_day <-
  divvy_trips %>%
    group_by(date = date(date)) %>%
    summarize (riders = n())

divvy_by_day$trip_identity = 'divvy'

# TAXI DEMAND GROUPED BY DAY
taxi_by_day <-
  taxi_start_times_date %>%
  filter(date != '2017-08-01') %>%
    group_by(date = date) %>%
    summarize (riders = n())

taxi_by_day$trip_identity = 'taxi'

# MERGED DATAFRAME OF TRANSPORTATION
total_demand_by_day <- bind_rows(public_transport_by_day,divvy_by_day, taxi_by_day)
remove(divvy_by_day, public_transport_by_day, taxi_by_day)
```

```{r, warning=FALSE, message=FALSE}
# PLOTTING GRAPH of 3 Year Chicago Public Transportation Trend + Weather


# WINTER
# Shading Winters From 21st December to 30th March
# Reference: https://stackoverflow.com/questions/18419628/creating-custom-legends-in-ggplot2

df<-data.frame(xmin=as.Date(c('2015-01-01','2015-12-21','2016-12-21', '2017-12-21' )),
               xmax=as.Date(c('2015-03-20','2016-03-30', '2017-03-30', '2017-12-31' )),
               ymin=c(-Inf,-Inf),
               ymax=c(Inf,Inf),
               winters=c("Winters"))


total_demand_by_day %>%
  ggplot(aes(x = date, y = riders/1000)) +
  geom_smooth(aes(color = trip_identity), se = FALSE, na.rm = TRUE) +
  geom_rect(data=df,aes(xmin=xmin,ymin=ymin,xmax=xmax,ymax=ymax, fill = 'winters'),
                    alpha=0.15,inherit.aes=FALSE, fill = winter_color) + 
  scale_color_manual(values=c(bus_color,divvy_color, taxi_color, train_color)) +
  scale_x_date(date_labels = "%d %b %Y") +
  theme_bw() +
  my_theme +
  theme(legend.position="none") +
  scale_y_continuous(labels = scales::comma) +
  geom_vline(xintercept = as.Date(c("2015-01-01", "2016-01-01", "2017-01-01", "2018-01-01")), 
             linetype = 4)  +
  annotate("label", x = as.Date("2015-06-01"), y = 500, 
           label = 'TRAIN', color = train_color, fontface =2,  family="Courier")  +
  annotate("label", x = as.Date("2016-06-01"), y = 670, 
           label = 'BUS', color = bus_color, fontface =2,  family="Courier") +
  annotate("label", x = as.Date("2016-06-01"), y = 100, 
           label = 'TAXI', color = taxi_color, fontface =2,  family="Courier") +
  annotate("label", x = as.Date("2015-05-01"), y = 40, 
           label = 'DIVVY', color = divvy_color, fontface =2,  family="Courier")  +
  annotate("label", x = as.Date("2016-03-21"), y = 320, 
           label = 'WINTER', color = winter_color, fontface =2,  family="Courier") +
  labs(title ="Demand for Public Transport & Divvy In Chicago Is Lowest In Winter", 
     subtitle = "Taxi Usage Less Elastic To Seasonal Changes",
     x = "", y = "Passengers Per Day (Thousands)",
     caption = "CTA Ridership Data\n 2015-2017\n *Taxi Data till July 2017")  +
  annotate("text", x = as.Date("2017-08-10"), y = 30, 
         label = '*', fontface =2,  family="Courier") 
```

Between 2015 - 2017, there are two very clear trends:
1) For Trains & Divvy bikes, there is a prominent seasonal variation. In the summer, demand peaks, while in the winter, demand is at its lowest.
2) Demand for Buses & Taxis has decreased throughout this time period. There also doesn't seem to be a seasonal component accompanying this decline.
These patterns can impact the allocation and deployment of transportation services by the CTA across seasons. One possible explanation for the decline of demand for buses and taxis might be an increased reliance on ride-share services such as Uber, Lyft etc.


<br><br>


```{r, warning=FALSE, message=FALSE}
temp <- 
  public_transport %>%
    group_by(dayofweek, trip_identity) %>%
    summarise(count = (sum(rides)/(3*52.19))/1000)

bus_point <-  
  temp %>%
    filter(trip_identity == "bus")

train_point <-  
  temp %>%
    filter(trip_identity == "train")
  
ggplot()  +
  geom_point(data = bus_point, aes(x = dayofweek, y = count), colour = bus_color, size = 5, stroke = 1) + 
  geom_point(data = train_point, aes(x = dayofweek, y = count), colour = train_color, size = 5, stroke = 1) + 
  geom_line(data = temp, mapping = aes(x = dayofweek, y=count),  alpha = 0.3)  +
  scale_y_continuous(labels = scales::comma) +
  coord_cartesian( ylim = c(200, 900)) + 
  labs(title ="Usage of Public Transport in Chicago is Highest on Weekdays", 
       subtitle = "Weekends see a significant drop in riders",
       x = "Day of Week", y = "Average Daily Passengers (Thousands)",
       caption = "CTA Ridership Data\n 2015-2017") +
  theme_bw()  + 
  theme(plot.title = element_text(hjust = -0, size = 12, face = "bold"), 
        plot.subtitle = element_text(size = 10, face = "bold"), 
        legend.title=element_blank(),
        legend.key = element_rect(fill = "white")) +
  coord_flip() +
  my_theme + 
  annotate("label", x = 5.5, y = 370, 
         label = 'TRAIN', color = train_color, fontface =2,  family="Courier") + 
  annotate("label", x = 5.5, y = 510, 
         label = 'BUS', color = bus_color, fontface =2,  family="Courier") + 
  theme(legend.position="none",
        axis.title.y=element_blank(),
        axis.ticks.y=element_blank()
        )
```

For buses and trains, comparisons within a week indicate that highest usage is during weekdays, possibly due to work or school transits. Deployments of public transportation are also more spread out on weekends by the CTA to ensure more efficient use.

<br><br>
```{r, warning=FALSE, message=FALSE}

taxi_num_days <- n_distinct(date(taxi_start_times_date$date))
divvy_num_days <- n_distinct(date(divvy_trips$date))

taxi_start_times_date$hour <- hour(taxi_start_times_date$date)
taxi_start_times_date$weekday <- lubridate::wday(taxi_start_times_date$date, label = TRUE)

taxi_by_weekday <-
  taxi_start_times_date %>%
    group_by(weekday) %>%
    summarise(count = n())

divvy_by_weekday <-
  divvy_trips %>%
  group_by(weekday = wday(divvy_trips$date)) %>%
  summarize(count = n())

taxi_by_weekday$weekday <- factor(taxi_by_weekday$weekday, levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun"))

ggplot()  +
  geom_point(data = taxi_by_weekday, aes(x = weekday, y = count/taxi_num_days), 
             colour = taxi_color, size = 5, stroke = 1) +
  geom_point(data = divvy_by_weekday, aes(x = weekday, y = count/divvy_num_days), 
             colour = divvy_color, size = 5, stroke = 1) +
  scale_y_continuous(labels = scales::comma) +
  #coord_cartesian( ylim = c(200, 900)) + 
  labs(title ="Taxi Demand Peaks Throughout Week, Then Falls On Weekend", 
       subtitle = "Divvy Bike Demand Remains Constant",
       x = "Day of Week", y = "Average Daily Trips",
       caption = "Chicago Taxi & Divvy \n Ridership Data 2015-2017") +
  theme_bw()  + 
  theme(plot.title = element_text(hjust = -0, size = 12, face = "bold"), 
        plot.subtitle = element_text(size = 10, face = "bold"), 
        legend.title=element_blank(),
        legend.key = element_rect(fill = "white")) +
  #coord_flip() +
  my_theme + 
  annotate("label", x = 5, y = 7700, 
         label = 'TAXI', color = taxi_color, fontface =2,  family="Courier") + 
  annotate("label", x = 5, y = 2210, 
         label = 'DIVVY', color = divvy_color, fontface =2,  family="Courier") + 
  theme(legend.position="none",
        axis.title.y=element_blank(),
        axis.ticks.y=element_blank()
        )

```

The first interesting observation is that demand for divvy doesn't vary much throughout the week, even though one might expect weekends to have higher demand, especially in downtown areas. The second interesting observation is that demand for taxis keeps increasing between Monday - Thursday even though there isn't any significant different in activities/behaviors of individuals between those days. Some of the increase in demand for Friday can be attributed to Friday-Night outgoing behavior. While weekends see a decline, Sunday is much lower than Saturday, which is also surprising.


<br><br>
```{r, warning=FALSE, message=FALSE}

winter_dates <- 
  weather %>%
  filter((date >= "2015-01-01" & date <= "2015-03-20") |
           (date >= "2015-12-21" & date <= "2016-03-20") |
           (date >= "2016-12-21" & date <= "2017-03-20") |
           (date >= "2017-12-21" & date <= "2017-12-31"))

snow_days <-
  winter_dates %>% 
  filter (SNOW > 0)

no_snow_days <-
  winter_dates %>% 
  filter (SNOW == 0)
  
train_snow_days <- 
  merge(snow_days, train_trips, by = "date") %>%
  group_by(date) %>%
  summarize (riders = sum(rides)) %>%
  mutate (type = "TRAIN", day_type = "SNOW")

train_no_snow <- 
  merge(no_snow_days, train_trips, by = "date") %>%
  group_by(date) %>%
  summarize (riders = sum(rides)) %>%
  mutate (type = "TRAIN", day_type = "NO SNOW")

bus_snow_days <- 
  merge(snow_days, bus_trips, by = "date") %>%
  group_by(date) %>%
  summarize (riders = sum(rides)) %>%
  mutate (type = "BUS", day_type = "SNOW")

bus_no_snow <- 
  merge(no_snow_days, bus_trips, by = "date") %>%
  group_by(date) %>%
  summarize (riders = sum(rides)) %>%
  mutate (type = "BUS", day_type = "NO SNOW")

snow_comparison <- do.call("rbind", list(train_snow_days, train_no_snow, bus_snow_days,bus_no_snow ))
snow_comparison$day_type <- factor(snow_comparison$day_type, levels = c("SNOW", "NO SNOW"))
remove (train_snow_days, train_no_snow, bus_snow_days,bus_no_snow, snow_days, no_snow_days)

ggplot(snow_comparison) +
  geom_boxplot(aes(x = day_type, y = riders/1000,  fill = type)) + 
  scale_x_discrete(labels = c('Rides Taken on Snow Days \n in Winter',
                              'Rides Taken on Non-Snow Days \n in Winter')) +
  labs(title ="Snow Does NOT Have Much Effect on Ridership in Winter", 
       subtitle = "Cold Non-Snowy Days Seem As Unappealing To Riders",
       x = "", 
       y = "Riders (Thousands)",
       caption = "NCEI & CTA Rider Data\n 2015-2017") +
  theme_bw() +
  my_theme +
  scale_fill_manual(values=c(bus_color, train_color)) +
  annotate("label", x = 0.8, y = 200, 
           label = 'BUS', color = bus_color, fontface =2,  family="Courier") +
  annotate("label", x = 1.8, y = 120, 
           label = 'BUS', color = bus_color, fontface =2,  family="Courier") +
  annotate("label", x = 1.2, y = 750, 
           label = 'TRAIN', color = train_color, fontface =2,  family="Courier")  +
  annotate("label", x = 2.2, y = 750, 
           label = 'TRAIN', color = train_color, fontface =2,  family="Courier") +
  theme(legend.position="none")
  
```

Comparing within cold days, it can be seen that consumers are indifferent between the presence & absence of snow. As such, snow doesn't necessarily present a physical barrier that would make people want to take fewer buses or trains (on average).

<br><br>

```{r, warning=FALSE, message=FALSE}
# reference: https://ibecav.github.io/slopegraph/

p <- public_transport %>%
  group_by(year, month, trip_identity) %>%
  summarise(riders = sum(rides)) %>%
  filter (trip_identity == "bus") 

green_palette <- c("#467563", "#78c679", "#78c679", "#238443","#467563", 
                   "#c2e699", "#78c679", "#238443","#78c679", "#78c679", "#78c679", "#238443")

alternative <- c(train_color, divvy_color, taxi_color, winter_color, 
                 "#525252", "#02818a", train_color, 
                 divvy_color, taxi_color, winter_color, "#e31a1c", "#6a51a3")

ggplot(p, aes(x = year, y = riders, group = month)) +
  scale_x_discrete(labels = c('Rides Taken on Snow Days \n in Winter',
                              'Rides Taken on Non-Snow Days \n in Winter', "wow"))  +
  geom_line(aes(color = month), alpha = 1, size = 1, linetype = 1) +
  geom_point(aes(color = month, alpha = 1), size = 1.5) + 
  scale_colour_manual(values=green_palette) +
  geom_label_repel(data = p %>% filter(year == "2015"),
                  aes(label = paste0(month), color = month, family = "Courier", fontface = 2) , 
                  hjust = "left", 
                  fontface = 7, 
                  size = 3, 
                  nudge_x = -.45, 
                  direction = "y") +
  geom_label_repel(data = p %>% filter(year == "2017"),
              aes(label = paste0(month), color = month, family = "Courier", fontface = 2) , 
              hjust = "right", 
              fontface = 7, 
              size = 2.5, 
              nudge_x = .4,
              direction = "y") +
  #scale_x_discrete(position = "bottom") +
  theme_bw() +
  theme(legend.position = "none") +
  theme(panel.border     = element_blank()) +
  #theme(axis.title.x     = element_blank()) +
  theme(axis.text.y     = element_blank()) +
  theme(axis.ticks   = element_blank()) +
    my_theme +
  labs( title = "Overall Bus Usage Has Declined Between 2015 and 2017",
    subtitle = "Summer Months still experience more bus usage than in Winter",
    caption = "CTA Transit Data\n 2015 - 2017",
    x = "2015                     2016                      2017",
    y = "Total Bus Riders") +
  theme(plot.title = element_text(hjust = 0.58, size = 12, face = "bold"), 
        plot.subtitle = element_text(hjust = 0.515, size = 10, face = "bold")) + 
    scale_y_continuous(labels = scales::comma) 

```

The demand for buses has been falling between 2015 and 2017 as established earlier. However, further exploration reveals that this decline isn't driver by any particular set of months or seasonality, but by an overall decline across each month. 

<br><br>

```{r, warning=FALSE, message=FALSE}

divvy_by_gender <- divvy_trips %>%
  filter(GENDER != '') %>%
  group_by(hour_of_day = hour(date), GENDER) %>%
  summarize(riders_by_hour = trunc(n() / (365*3)))


ggplot(data = divvy_by_gender) +
  geom_col(data = subset(divvy_by_gender, GENDER == "Male"), 
           aes(x = hour_of_day, y = riders_by_hour, fill = GENDER), fill = divvy_male_color) +
  geom_col(data = subset(divvy_by_gender, GENDER == "Female"), 
           aes(x = hour_of_day, y = -riders_by_hour, fill = GENDER), fill = divvy_female_color) + 
  scale_fill_brewer(palette = "Set1") + 
  theme_bw() +
  coord_flip() +
  geom_hline(yintercept=0, colour="white", lwd=1) + my_theme + 
  scale_y_continuous( labels = abs) +
  annotate("label", x = 23.5, y = -140,
           label = 'FEMALE RIDERS', color = divvy_female_color, fontface =2,  family="Courier") +
  annotate("label", x = 23.50, y = 250, 
           label = 'MALE RIDERS', color = divvy_male_color, fontface =2,  family="Courier") +
  labs(title ="Highest Demand For Divvy Bikes During Peak Morning & Evening Hours", 
     subtitle = "Significantly Fewer Female Riders, Especially At Night",
     x = "", y = "Passengers Per Day",
     caption = "Divvy Ridership Data\n 2015-2017") + 
  scale_x_continuous(sec.axis = dup_axis(), breaks = c(0,6,12,17,22), 
                     labels = c("Midnight", "6 am", "Noon", "5 pm", "10 pm" )) +
  theme(axis.title.x = element_text(hjust = 0.2))

#remove(divvy_by_gender)
```

Demand for divvy bikes peaks at morning and evening rush hours. There is a noticably higher usage of divvy bikes by men compared to women at all hours of the day. It's also interesting to note that there are almost no women riding bikes between 1 am and 4 am.


<br><br>

```{r, warning=FALSE, message=FALSE}

divvy_by_duration <-
divvy_trips %>%
  group_by(day_of_week, time_of_day) %>%
  summarize(duration = median(TRIP.DURATION)) 

divvy_by_duration$day_of_week <- factor(divvy_by_duration$day_of_week, levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun"))


divvy_by_duration$time_of_day <- factor(divvy_by_duration$time_of_day, 
                            levels = rev(c("0", "1","2", "3","4", "5","6", "7","8", "9","10", 
                                           "11","12", "13","14", "15","16", 
                                       "17","18", "19",
                                       "20", "21","22", "23")))


ggplot(data = divvy_by_duration, aes(x = day_of_week, y = time_of_day, fill = duration)) +
geom_tile()+
  scale_fill_gradient(low = "#bdd7e7", high = "#2171b5" ,
                      name = "Median Trip\nDuration\n(In Seconds)", labels = comma) +
  theme_bw() +
  my_theme +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +  
  theme(axis.title.x=element_blank(),
        axis.title.y=element_blank(),
        axis.ticks.y=element_blank(),
        axis.ticks.x=element_blank(),
        axis.text.y = element_text(vjust=-1.3),
        axis.text.x = element_text(size = 8, face = "bold")) +
  theme (legend.key = element_rect(fill = "black"),
        legend.spacing.x = unit(0.2, 'cm'),
        legend.text = element_text(family = "Courier", size = 8, face = "bold"),
        legend.title = element_text(family = "Courier", size = 8, face = "bold"),
        legend.background = element_rect(fill="white", size=1, linetype="solid",colour ="#969696")) +
  scale_x_discrete (position = "top", expand = c(0,0)) +
  scale_y_discrete(breaks = c(0,3,6,9,12,15,18,21),
                   labels = c("Midnight","3 am","6 am","9 am", "Noon", "3 pm", "6 pm", "9 pm"))  +
  labs (title ="Longest Divvy Trips Occur on Weekends.", 
       subtitle = "Weekdays See Shorter Rides, Especially in Mornings & Evenings.",
       x = "", 
       y = "",
       caption = "Chicago Divvy Data\n 2015-2017") 

```

Possibly due to increased tourist and visitor usage in Downtown/River area on weekends, the median trip duration is quite large compared to weekday median trip durations. Late night trips also seem to be of large durations, which may most likely be attributed to a very low number of riders in the first place, who may be interested in travelling for long durations at that time of day.


<br><br>
```{r, warning=FALSE, message=FALSE}

taxi_start_times_solo <- as.data.frame(taxi_start_times)

taxi_light <- "#ffffb2"
taxi_dark <- "#cc4c02"

taxi_start_times_solo$hour <- hour(taxi_start_times_solo$taxi_start_times)
taxi_start_times_solo$weekday <- lubridate::wday(taxi_start_times_solo$taxi_start_times, label = TRUE)

taxi_by_hour <-
  taxi_start_times_solo %>%
  group_by(weekday, hour) %>%
  summarize(count = n())

taxi_by_hour$weekday <- factor(taxi_by_hour$weekday, levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun"))
taxi_by_hour$hour <- as.factor (taxi_by_hour$hour)
taxi_by_hour$hour <- factor(taxi_by_hour$hour, 
                            levels = rev(c("0", "1","2", "3","4", "5","6", "7","8", "9","10", 
                                           "11","12", "13","14", "15","16", 
                                       "17","18", "19",
                                       "20", "21","22", "23")))


ggplot(data = taxi_by_hour, aes(x = weekday, y = hour,  fill = count/taxi_num_days)) +
  geom_tile() +
  scale_fill_gradient(low = taxi_light, high = taxi_dark,name = "Taxi Trips", labels = comma) +
  theme_bw() +
  my_theme + 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +  
  theme(axis.title.x=element_blank(),
        axis.title.y=element_blank(),
        axis.ticks.y=element_blank(),
        axis.ticks.x=element_blank(),
        axis.text.y = element_text(vjust=-1.3),
        axis.text.x = element_text(size = 8, face = "bold")) +
  theme (legend.key = element_rect(fill = "black"),
        legend.spacing.x = unit(0.2, 'cm'),
        legend.text = element_text(family = "Courier", size = 8, face = "bold"),
        legend.title = element_text(family = "Courier", size = 8, face = "bold"),
        legend.background = element_rect(fill="white", size=1, linetype="solid",colour ="#969696")) +
  scale_x_discrete (position = "top", expand = c(0,0)) +
  scale_y_discrete(breaks = c(0,3,6,9,12,15,18,21),
                   labels = c("Midnight","3 am","6 am","9 am", "Noon", "3 pm", "6 pm", "9 pm"))  +
  labs (title ="Highest Weekday Taxi Traffic Between 6-8 pm, Much Later on Weekends", 
       subtitle = "Weekend Rush Drives Taxis Late Into The Night",
       x = "", 
       y = "",
       caption = "Chicago Taxi Trips Data\n 2015-2017") 
```

During weekdays, high demand is during the day, peaking during evening rush hour, with Friday evening rush hour being the busiest. However, unlike on weekdays, on weekdays, demand stays high at night, and extends to the early hours of the next day on Saturday and Sunday. 
<br><br>

#### NEXT STEPS:
Next steps involve creating maps & vizualizations that highlight flow of traffic & changes in demand patterns & routes.



```{r}

taxi_2015 <- read.csv (here::here("data", "2015_Taxi_Trips1.csv"))
taxi_2015 <-
  taxi_2015 %>%
  drop_na(Pickup.Centroid.Latitude, Pickup.Centroid.Longitude, Pickup.Centroid.Location,
          Dropoff.Centroid.Latitude, Dropoff.Centroid.Longitude, Dropoff.Centroid..Location,
          Trip.Total)


taxi_2015$Pickup.Centroid.Latitude = as.numeric(as.character(taxi_2015$Pickup.Centroid.Latitude))
taxi_2015$Dropoff.Centroid.Latitude = as.numeric(as.character(taxi_2015$Dropoff.Centroid.Latitude))
taxi_2015$Dropoff.Centroid.Longitude = as.numeric(taxi_2015$Dropoff.Centroid.Longitude)


taxi_2015 <- 
  taxi_2015 %>%
  drop_na(Pickup.Centroid.Latitude) %>%
  mutate_if(is.numeric, round, digits=4)


taxi_2015_grouped <-
  taxi_2015 %>%
    group_by(Pickup.Centroid.Longitude, Pickup.Centroid.Latitude) %>%
    summarize(median_cost = median(Trip.Total), num_trips = n()) %>%
    filter (median_cost < 100, num_trips > 10)

```



```{r}
my_theme_maps <- 
  theme(
    axis.text = element_text(size = 7, face = "bold", family="Courier"),
    axis.title =  element_text(size = 10, face = "bold", family="Courier"),
    plot.title = element_text(hjust = 0, size = 14, face = "bold", family="Courier"), 
    plot.subtitle = element_text(size = 12, face = "bold", family="Courier"),
    plot.caption = element_text(size = 8, face = "bold", family="Courier"),
    legend.spacing.x = unit(0.2, 'cm'),
    panel.border = element_rect(colour = "#969696", fill=NA, size=2),
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank(),
    axis.title.x=element_blank(),
    axis.text.x=element_blank(),
    axis.ticks.x=element_blank(),
    axis.title.y=element_blank(),
    axis.text.y=element_blank(),
    axis.ticks.y=element_blank(),
    legend.text = element_text(family = "Courier", size = 8, face = "bold"),
    legend.title = element_text(family = "Courier", size = 8, face = "bold"),
    legend.background = element_rect(fill="white", size=1, linetype="solid",colour ="#969696"))

```




```{r}


taxi_starts <- st_as_sf(taxi_2015_grouped, coords=c("Pickup.Centroid.Longitude","Pickup.Centroid.Latitude"),
                        agr="constant", crs=4326)

areas <- st_read ("https://data.cityofchicago.org/resource/igwz-8jzy.geojson")
areas <- st_transform (areas, 32616)


ggplot(data = taxi_starts) +
  geom_sf(data = areas, colour = "white", fill = "black") +
  geom_sf(aes(size = num_trips, colour = median_cost),  show.legend = 'point') +
  theme_bw() +
  my_theme_maps +
  theme(panel.grid.major=element_line(colour="transparent")) +
  scale_colour_gradient(low = "#E8E013", high = "#FF1570",
                        name = "Median Trip Cost\n(In USD)", labels = comma) +
  annotate("segment", x = -87.91, xend = -87.92, y = 41.97, yend = 41.93, 
           color="orange", linetype="twodash") +
  annotate("text", x = -87.91, y = 41.92, label = "O'Hare \nAirport", 
         colour = "orange", face = "bold", size = 3, face = "bold", family="Courier" ) +
  annotate("segment", x = -87.765, xend = -87.83, y = 41.785, yend = 41.82, 
           color="orange", linetype="twodash") +
  annotate("text", x = -87.83, y = 41.83, label = "Midway \nAirport", 
         colour = "orange", face = "bold", size = 3, face = "bold", family="Courier" ) +
  annotate("segment", x = -87.615, xend = -87.58, y = 41.895, yend = 41.93, 
         color="orange", linetype="twodash") +
  annotate("text", x = -87.57, y = 41.94, label = "Downtown \nChicago", 
         colour = "orange", face = "bold", size = 3, face = "bold", family="Courier" ) +
  #theme(legend.position = c(0.13,0.24))  +
  scale_size_continuous(name = 'Number of Rides \n(From Location)', labels = comma) +
  labs (title ="Airports & South Side Pickups Generate Highest Median Revenue for Taxi Drivers", 
       subtitle = "But Most Taxi Traffic Is Concentrated in Downtown & North Side",
       x = "", 
       y = "",
       caption = "Chicago Taxi Trips\n2015 Data") 



```





```{r}

taxi_2015_grouped_distance <-
  taxi_2015 %>%
    group_by(Pickup.Centroid.Longitude, Pickup.Centroid.Latitude) %>%
    summarize(median_trip_distance = median(Trip.Miles), num_trips = n()) %>%
    filter (median_trip_distance > 0.06, num_trips > 10)

taxi_starts_distance <- st_as_sf(taxi_2015_grouped_distance,
                                 coords=c("Pickup.Centroid.Longitude","Pickup.Centroid.Latitude"),
                                 agr="constant", crs=4326)


ggplot(data = taxi_starts_distance) +
  geom_sf(data = areas, colour = "white", fill = "black") +
  geom_sf(aes(size = num_trips, colour = median_trip_distance),  show.legend = 'point') +
  theme_bw() +
  my_theme_maps +
  theme(panel.grid.major=element_line(colour="transparent")) +
  scale_colour_gradient(low = "#E8E013", high = "#FF1570",
                        name = "Median Trip \nDistance\n(In Miles)", labels = comma) +
  annotate("segment", x = -87.91, xend = -87.92, y = 41.97, yend = 41.93, 
           color="orange", linetype="twodash") +
  annotate("text", x = -87.91, y = 41.92, label = "O'Hare \nAirport", 
         colour = "orange", face = "bold", size = 3, face = "bold", family="Courier" ) +
  annotate("segment", x = -87.765, xend = -87.83, y = 41.785, yend = 41.82, 
           color="orange", linetype="twodash") +
  annotate("text", x = -87.83, y = 41.83, label = "Midway \nAirport", 
         colour = "orange", face = "bold", size = 3, face = "bold", family="Courier" ) +
  annotate("segment", x = -87.615, xend = -87.58, y = 41.895, yend = 41.93, 
         color="orange", linetype="twodash") +
  annotate("text", x = -87.57, y = 41.94, label = "Downtown \nChicago", 
         colour = "orange", face = "bold", size = 3, face = "bold", family="Courier" ) +
  #theme(legend.position = c(0.13,0.24))  +
  scale_size_continuous(name = 'Number of Rides \n(From Location)', labels = comma) +
  labs (subtitle ="Downtown has Shortest Rides", 
       title = "Excluding Airports, Highest Number of Long Distance Taxi Rides Are From Hyde Park. ",
       x = "", 
       y = "",
       caption = "Chicago Taxi Trips\n2015 Data") 

```

#### OKAY NEXT



```{r}

areas <- st_read ("https://data.cityofchicago.org/resource/igwz-8jzy.geojson")
areas <- st_transform (areas, 32616)


trains <- st_read ("/Users/kazi/Desktop/transportation-demand-chicago/data/CTA_RailLines/CTA_RailLines.shp ")
trains <- st_read(here::here("data/CTA_RailLines/", "CTA_RailLines.shp"))
trains <- st_transform (trains, 32616)




ggplot() +
  geom_sf(data = areas, colour = "white", fill = "white") +
  geom_sf(data = trains, colour = "red")
```



```{r}

divvy_trips <-
  divvy_trips %>%
  drop_na(FROM.LONGITUDE,FROM.LATITUDE)

divvy_from_sf <- st_as_sf(divvy_trips, 
                          coords=c("FROM.LONGITUDE","FROM.LATITUDE"), 
                          crs=4326, 
                          agr="constant")



ggplot() +
  geom_sf(data = areas) +
  geom_sf (data = water_wards, aes(fill = ))

intersects <- st_intersects(areas, mini_divvy)
water_wards <- st_join(mini_divvy, areas)

table(intersects$id_polygons)

areas <- st_read ("https://data.cityofchicago.org/resource/igwz-8jzy.geojson")
areas <- st_transform (areas, 32616)


mini_divvy <- st_transform (mini_divvy, 32616)

st_crs(mini_divvy)

x<- water_wards %>%
  group_by(community) %>%
  summarize (bike_starts = n()) 

  ggplot() +    
  geom_sf(data = areas) +
  geom_sf(data = x, aes(fill = bike_starts))

plot(areas)

```






```{r}

divvy_trips <-
  divvy_trips %>%
  drop_na(FROM.LONGITUDE,FROM.LATITUDE)

divvy_small <- st_as_sf(divvy_trips, coords=c("FROM.LONGITUDE","FROM.LATITUDE"),
                        agr="constant", crs=4326)

divvy_small <- st_transform (divvy_small, 32616)
st_contains(divvy_small, areas)
```



```{r}

#1 week before week of lol
#2 week == effect of week of lol
#1 week after effect of week of lol


downtown_effected_buses <- c("3", "4", "12", "6", "10", "J14", "130", "146",
                             "126", "147", "29", "62", "36", "151", "60", "22")
#2015
lol_2015 <-
  bus_trips  %>%
  filter(date >= "2015-07-20" & date <= "2015-08-16") %>%
  select(route, date, rides) 
lol_2015$days_to_lolla <- lol_2015$date - date("2015-07-31")

#2016
lol_2016 <-
  bus_trips  %>%
  filter(date >= "2016-07-18" & date <= "2016-08-14") %>%
  select(route, date, rides) 
lol_2016$days_to_lolla <- lol_2016$date - date("2016-07-28")

#2017
lol_2017 <-
  bus_trips  %>%
  filter(date >= "2017-07-24" & date <= "2017-08-20") %>%
  select(route, date, rides) 
lol_2017$days_to_lolla <- lol_2017$date - date("2017-08-03")

lol_total <- rbind(lol_2015, lol_2016, lol_2017)

lol_effected <- 
  lol_total %>% 
  filter(route %in% downtown_effected_buses)  %>%
  group_by(days_to_lolla) %>%
  #summarise(riders = mean(rides))
  summarise(riders = sum(rides))

lol_effected = lol_effected[-1,]
lol_effected <- lol_effected[-nrow(lol_effected),]

lol_uneffected <- 
  lol_total %>% 
  filter(!(route %in% downtown_effected_buses)) %>%
  group_by(days_to_lolla) %>%
  #summarise(riders = mean(rides)) 
  summarise(riders = sum(rides))

lol_uneffected = lol_uneffected[-1,]
lol_uneffected <- lol_uneffected[-nrow(lol_uneffected),]

ymin=c(-Inf,-Inf)
ymax=c(Inf,Inf)

bus_downtown <- "#66c2a4"
bus_other <- "#2ca25f"
lolla_color <- "#f03b20"


img <- readPNG(here::here("lolla_image.png"))
g <- rasterGrob(img, interpolate=TRUE)

ggplot() +
  stat_smooth(data = lol_effected, aes(x = days_to_lolla, y = riders/(3*1000)), 
              colour = bus_downtown, method = "lm",
              formula = y ~ poly(x, 15), se = FALSE)  +
  stat_smooth(data = lol_uneffected, aes(x = days_to_lolla, y = riders/(3*1000)), 
              colour = bus_other, method = "lm",
              formula = y ~ poly(x, 15), se = FALSE)  +
  scale_y_continuous(labels = scales::comma) +
  scale_x_continuous(breaks = c(-14:20)) +
  theme_bw() +
  my_theme +
  annotation_custom(g, xmin=0, xmax=4, ymin=700, ymax=800) +
  scale_x_continuous(breaks = c(-10,-3,0,4,11, 14),
                   labels = c("10 days To\nLollapalooza","Road\nBlocks\nStart", 
                              "Start Of\nLollapalooza\n", 
                              "End of\nLollapalooza", "Road\nBlocks\nEnd",
                              "10 Days After\nEnd Of\nLollapalooza")) +
  theme(axis.title.x=element_blank()) +
  annotate("label", x = 13, y = 680, 
           label = 'Route-Unaffected\nBuses', color = bus_other, 
           fontface =2,  family="Courier") +
  annotate("label", x = 12, y = 220, 
           label = 'Route-Affected Buses', color = bus_downtown, 
           fontface =2,  family="Courier")   + 
  annotate("text", x = 1, y = 320, label = "Higher Trough\nThan Usual", 
         colour = lolla_color, face = "bold", size = 2.5, face = "bold", family="Courier" ) +
  annotate("segment", xend = 2, x = 1, yend = 400, y = 350, 
         color=lolla_color,
         arrow = arrow(length = unit(0.35, "cm")))  +
  labs(title ="Regular Users Switch to Route-Unaffected Buses During Lollapalooza.", 
       subtitle = "Users Revert to Regular Demand Patterns After Road Closures Removed.",
       x = "", y = "Bus Users (Thousands)",
       caption = "CTA Bus Data\n 2015-2017") 
  
```






